---
title: 'Getting Started with R/Rstudio'
author: "Michael McIsaac"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

### Load a few helpful packages (after first installing them if they aren't already installed). 

#specify the packages of interest
packages = c("api2lm", "mosaic", "car", "leaps", "tidyverse", "here")

#use this function to check if each package is on the local machine
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed and loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(char=packages)
```

## Read in the data
In order to read the data into R, you need to enter the path to the dataset. 

You should create a new "Project" in RStudio and, once you download your .csv file, you should place that dataset inside your project folder (ideally in a sub-folder called something like "data").
That way, your path could be relative to project directory. 
Alternately, you could use something like `file.choose()` to bring up a GUI that allows you to point and click on the file, but this means you need to point and click everytime you re-run the code and your code would not be reproducible (someone else couldn't run your Rmd file and get the same output). 

```{r read_in_data}
# Now read in the data from the defined path:
senic = read_csv(here("data", "senic.csv"))
```
## Regression Modelling 

Run a linear model based on the data set that we called "senic" and summarize the model fit

```{r}
model_infection = lm(length_of_stay~infection_risk, data=senic)
msummary(model_infection)
```

Find the confidence intervals for the parameters in the model

```{r}
confint(model_infection)
```

Find the pointwise confidence intervals for a few specific predictions

```{r}
predict(model_infection, newdata=data.frame(infection_risk=c(5, 5.5, 6)), interval="confidence", level=.95)
```

Find the pointwise prediction intervals for a few specific predictions
```{r}
predict(model_infection, newdata=data.frame(infection_risk=c(5, 5.5, 6)), interval="prediction", level=.95)
```

Find the Working-Hotelling confidence band values for a few specific predictions

```{r}
predict_adjust(model_infection, newdata=data.frame(infection_risk=c(5, 5.5, 6)), interval="confidence", method="wh", level=.95)
```


Find the Bonferroni simultaneous confidence interval values for a few specific predictions

```{r}
predict_adjust(model_infection, newdata=data.frame(infection_risk=c(5, 5.5, 6)), interval="confidence", method="bonferroni", level=.95)
```

Run a linear model based on the data set that we called "senic", but only the parts where "region" has the value "NC"

```{r}
model_infection_nc = lm(length_of_stay~infection_risk, data=senic, subset=(region=="NC"))
msummary(model_infection_nc)
```

Run a linear model based on the data set that we called "senic", but with the "infection_risk" variable centered at 0:

```{r}
model_infection_centered = lm(length_of_stay~I(infection_risk-mean(infection_risk)), data=senic)
msummary(model_infection_centered)
```

Run a linear model based on the data set that we called "senic", but with both "infection_risk" and "age" as predictor variables
```{r}
model_infection_age = lm(length_of_stay~infection_risk + age, data=senic)
msummary(model_infection_age)
```

Find the confidence interval for a specific prediction from the multiple linear regression model (infection risk is 5, age is 50): 

```{r}
predict(model_infection_age, newdata=data.frame(infection_risk=5, age=50), interval="confidence", level=.95)
```
Find the pointwise prediction intervals for a few specific prediction from the multiple linear regression model (1: infection risk is 5 and age is 50; 2: infection risk is 6 and age is 40): 

```{r}
predict(model_infection, newdata=data.frame(infection_risk=c(5, 6), age=c(50, 40)), interval="prediction", level=.95)
```




## Other potentially useful code and functions for future assignments

Create an ANOVA table
```{r}
anova(model_infection)
```

Find R^2

```{r}
summary(model_infection)$r.squared
```

Calculate Pearson correlation
```{r}
mosaic::cor.test(length_of_stay~infection_risk, data=senic, method="pearson")
```

Calculate Spearman correlation
```{r}
mosaic::cor.test(length_of_stay~infection_risk, data=senic, method="spearman", exact=FALSE)
```

Conduct a General Linear Test
```{r}
model_infection_reduced = lm(length_of_stay~1, data=senic)
anova(model_infection_reduced, model_infection)
```

Test for Lack of Fit

```{r}
pacman::p_load_gh("cran/alr3")
model_infection_full = lm(length_of_stay~factor(infection_risk), data=senic)
anova(model_infection, model_infection_full)
alr3::pureErrorAnova(model_infection)
```

Create Residual plots

```{r}
xyplot(length_of_stay~infection_risk, data=senic,  type=c("p", "smooth"))
xyplot(resid(model_infection)~senic$infection_risk, ylab="Residuals", xlab="Infection risk", type=c("p", "smooth"))
xyplot(resid(model_infection)~predict(model_infection), ylab="Residuals", xlab="Fitted Values", type=c("p", "smooth"))
xyplot(abs(resid(model_infection))~predict(model_infection), ylab="Absolute Residuals", xlab="Fitted Values", type=c("p", "smooth"))
xyplot(resid(model_infection)~senic$routine_culturing_ratio, ylab="Residuals", type=c("p", "smooth"))
boxplot(resid(model_infection), ylab="Residuals", horizontal=TRUE)
qqPlot(resid(model_infection), ylab="Residuals")
```

Box Cox Plot

```{r}
MASS::boxcox(model_infection)
```

LOWESS curve

```{r}
scatter.smooth(senic$length_of_stay, senic$infection_risk)
```

5 number summary

```{r}
summary(senic$infection_risk)
```

Change the order of a categorical (factor) variable (note that the first category listed is the reference category)
```{r}
senic$region = factor(senic$region, levels=c("W", "NE", "NC", "S"))
```



## Below is code that will likely be useful in the wrap-up assignment for Chapter 9:

Remove unwanted predictors
```{r}
senic = senic[, -c(7,8)]
```

Transform the outcome
```{r}
senic$length_of_stay = log10(senic$length_of_stay)
```

Separate the data set into a "training" set and a "validation" set
```{r}
train = senic[41:110, ]
validation = senic[1:40, ]
```

Set up a first-order linear regression model with all predictor variables
```{r}
stay_model = lm(length_of_stay~., data=train)
```

Calculate various model selection critiria for all possible models; retain only the 3 best models of each size (i.e., for each number of predictor variables)
```{r}
subsets = regsubsets(formula(stay_model), data=model.frame(stay_model), nvmax=10, nbest= 3)
```

Produce a data frame that displays the model selection criteria in an easier to read manner
```{r}
model_selection_criteria = data.frame(with(summary(subsets), cbind(p = as.numeric(rownames(which)), which, rss, rsq, adjr2, cp, bic)))
```

Order the models based on different criteria
```{r}
model_selection_criteria[order(model_selection_criteria$bic),]
model_selection_criteria[order(model_selection_criteria$rsq),]
model_selection_criteria[order(model_selection_criteria$adjr2),]
model_selection_criteria[order(model_selection_criteria$cp),]
```

Estimate parameters using the selected model(s). E.g., 

```{r}
model_best 			= lm(length_of_stay~infection_risk+routine_xray_ratio+average_daily_census, data=train)
model_secondbest= lm(length_of_stay~infection_risk+routine_xray_ratio+average_daily_census+number_of_nurses, data=train)
model_thirdbest	= lm(length_of_stay~infection_risk+routine_xray_ratio+number_of_beds, data=train)
```

Extract parameter estimates
```{r}
msummary(model_best)
msummary(model_best)$sigma^2
msummary(model_best)$r.squared
msummary(model_best)$adj.r.squared

msummary(model_secondbest)
msummary(model_thirdbest)
```

Estimate parameters for the best model(s) using the validation data set

```{r}
model_best.validation 			= lm(length_of_stay~infection_risk+routine_xray_ratio+average_daily_census, data=validation)

msummary(model_best.validation)
msummary(model_best.validation)$sigma^2
msummary(model_best.validation)$r.squared
msummary(model_best.validation)$adj.r.squared
```

Predict values for the validation data using the training model
```{r}
predict_fromtraining = model_best %>% predict(validation)
```

Predict values for the validation data using the validation model
```{r}
predict_withinvalidation = model_best.validation %>% predict(validation)

MSPR = mean((predict_fromtraining - validation$length_of_stay)^2)
```

## Below is code that will likely be useful in the wrap-up assignment for Chapter 10:


Note: you must define n and p appropriately for your data set:
```{r}
dim(senic)
n=100
p=4
```


```{r}
cor(senic[,c(2, 5, 9)])

model = lm(I(log10(senic$length_of_stay))~age+routine_xray_ratio+average_daily_census, data=senic)
vif(model)
sort(rstudent(model)) 
sort(hatvalues(model))

cooks.distance(model)
dffits(model)
dfbetas(model)

qt(1-.05/(2*n), df=n-p-1)
qf(.5, p, n-p)

```