---
title: 'Recap: Chapter 6-8'
subtitle: 'STAT 3240'
author: "Michael McIsaac"
institute: "UPEI"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false # disable slide transitions by scrolling
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
duo_accent(primary_color = "#006747", secondary_color = "#CFC493",   
	header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Droid Mono"))
```		



```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina = 3, warning = FALSE, message = FALSE, fig.align="center", fig.height=5.5)
options(DT.options = list(scrollX = TRUE, pageLength=20, scrollY = 250))
options(digits = 4)
options(show.signif.stars=FALSE)

library(here)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(tidyverse)
library(ggplot2)
library(knitr)
library(mosaic)
library(DT)
library(car)
library(psych)

#params
spending_subset_all = read.csv(here("data", "spending_subset.csv"))

spending_subset=spending_subset_all[1:500,]

par(lwd=3,cex=1.5) 
cdi = as_tibble(read.delim(here("data", "CDI.txt"), sep=" ", header=FALSE)[,-c(1:2)] %>% mutate(V18 = recode_factor(V18, "NE", "NC", "S", "W")))
names(cdi) = c("county", "state", "land_area", "population", "pop_18_to_34", "pop_65", "number_physicians", "number_hospital_beds", "total_serious_crimes", "high_school_grads", "bachelor_degrees", "poverty_rate", "unemployment_rate", "per_capita_income", "total_personal_income", "region")

tab_model <- function(...,  show.ci=.95){sjPlot::tab_model(...,  show.ci=show.ci, show.se=TRUE, collapse.ci=TRUE, show.stat=TRUE)}
```


### Learning Objectives for Chapter 6
You should be able to 

- Understand the concept and utility of multiple linear regression
- Interpret general linear regression coefficients
- Be aware of qualitative predictors, polynomial regression, and interactions

- Express model, estimation, fitted values, residuals, and ANOVA in matrix form
- Conduct and interpret a general linear regression ANOVA F test
- Calculate and interpret multiple $R^2$ and $r$
- Conduct and interpret inference and joint inference	on specific parameters

- Compute and interpret independent and simultatneous CIs for $E[Y_h]$ and PIs for new observations
- Apply regression diagnostics to the multiple regression setting. 

---

### Learning Objectives for Chapter 7
You should be able to 

- Understand the concept of the extra sums of squares principle
- Conduct and interpret tests concerning regression coefficients using ESS principle 

- Compute and interpret coefficients of partial determination
- Understand multicollinearity and its effects

---

### Learning Objectives for Chapter 8
You should be able to 

- Understand the utility and disadvantages of polynomial regression
- Understand the need for centering
- Understand the danger of overfitting
- Compute and interpret parameters in a polynomial regression model
- Understand the utility and disadvantages of interactions in regression
- Compute and interpret parameters in regression models with interactions
- Compute and interpret parameters in curvilinear regression models with interactions

- Implement and interpret regression using indicator (dummy) variables 

- Implement and interpret regression involving interactions between indicator and quantitative variables
- Implement and interpret regression involving interactions between multiple indicator variables
- Implement and interpret tests for differences among regression functions

---

### Test 2: Chapter 6-8 example

```{r}
spending_subset = spending_subset %>% select(income, sex, food_expenditure, clothing_expenditure)
clothing_model = lm(clothing_expenditure~income+sex+food_expenditure, data=spending_subset)
msummary(clothing_model)
```

---

```{r}
anova(clothing_model)
```

```{r}
round(confint(clothing_model, level=.95), 4)
```

---

```{r, echo=FALSE, fig.height=9, fig.width=9}
spending_subset$fitted_values = predict(clothing_model)
spending_subset$residuals = MASS::studres(clothing_model)


pairs.panels(spending_subset, 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = FALSE # show correlation ellipses
             )


```

---

```{r}
msummary(lm(clothing_expenditure~sex, data=spending_subset))
anova(lm(clothing_expenditure~sex, data=spending_subset))
```

---

```{r}
msummary(lm(clothing_expenditure~sex+(income + I(income^2))*sex, data=spending_subset))
anova(lm(clothing_expenditure~sex+(income + I(income^2))*sex, data=spending_subset))
```
